% run this through PhD-TeX

\input dcustom
\articlestyle
\input version

\def\appendix#1#2{\dosection{Appendix #1: #2}}

\def\draftext{%
    \begingroup
	\eightpoint\sf
	Reprinted from
	{\it Proceedings, Second International Symposium on Computer Message
	    Systems},
	{\oldstyle 1985}%
    \endgroup
}
\catcode`\@=11
\def\draftstring{%
    \ifnum\pageno>\z@
	\begingroup
	    \eightpoint\sf
	    Copyright $\copyright$ {\oldstyle 1985},
	    IFIP TC-6
	\endgroup
    \fi
}
\def\uci@footpage{\draftstring\hfil\rm\folio}
\catcode`\@=12

\input sfwmac
\def\CKD/{{\sf CKD}}
\def\CSM/{{\sf CSM}}
\def\KDC/{{\sf KDC}}
\def\KDS/{{\sf KDS}}
\def\MTA/{{\sf MTA}}
\def\MTS/{{\sf MTS}}
\def\TMA/{{\sf TMA}}
\def\TTI/{{\sl TTI\/}}
\def\UA/{{\sf UA}}

\def\itemm$#1${\item{$#1\/$:}}

\header
    Design of the TTI Prototype\\
	Trusted Mail Agent\title
    Marshall T.~Rose$^\dagger$\\
	David J.~Farber\\
	Stephen T.~Walker\author\info
\footnote{}{\hskip -\parindent
$^\dagger$ All three authors are with Trusted Technologies, Incorporated,
POB~45, Glenwood,~MD~21738, USA.
Telephone:	301/854--6889.
In addition, Professor Farber is with the University of Delaware.}

\centerline{\sc Abstract}
{\rightskip=0pt\narrower\narrower
\noindent
The design of the \TTI/ prototype Trusted Mail Agent (\TMA/) is discussed.
This agent interfaces between two entities:
a key distribution center (\KDC/) and a user agent (\UA/).
The \KDC/ manages keys for the encryption of text messages,
which two subscribers to a key distribution service (\KDS/) may exchange.
The \TMA/ is independent of any underlying message transport system.

\lp
Subscribers to the \KDC/ are known by unique identifiers,
known as IDs.
In addition to distributing keys,
the \KDC/ also offers a simple directory lookup service,
in which the ``real-world'' name of a subscriber may be mapped to an ID,
or the inverse mapping may be performed.

\lp
This document details three software components:
\underbar{first},
a prototype key distribution service,
which has been running in a TCP/IP environment since December, 1984;
\underbar{second},
a prototype trusted mail agent;
and, 
\underbar{third},
modifications to an existing \UA/,
the Rand \MH/ Message Handling system,
which permit interaction with the prototype \TMA/.
\par}

\input text
\input refs
\input appendixA
\input appendixB
\input appendixC
\printcontents
\unskip\footnote{}{\hskip -\parskip This document (version \versiontag/)
was \TeX set \today\ with DISS.STY v\version.}
\showsummary

\bye
