% $Id: title.tex,v 5.2 90/06/23 22:21:48 jsp Rel $
%
% Copyright (c) 1989 Jan-Simon Pendry
% Copyright (c) 1989 Imperial College of Science, Technology & Medicine
% Copyright (c) 1989 The Regents of the University of California.
% All rights reserved.
%
% This code is derived from software contributed to Berkeley by
% Jan-Simon Pendry at Imperial College, London.
%
% Redistribution and use in source and binary forms are permitted provided
% that: (1) source distributions retain this entire copyright notice and
% comment, and (2) distributions including binaries display the following
% acknowledgement:  ``This product includes software developed by the
% University of California, Berkeley and its contributors'' in the
% documentation or other materials provided with the distribution and in
% all advertising materials mentioning features or use of this software.
% Neither the name of the University nor the names of its contributors may
% be used to endorse or promote products derived from this software without
% specific prior written permission.
% THIS SOFTWARE IS PROVIDED ``AS IS'' AND WITHOUT ANY EXPRESS OR IMPLIED
% WARRANTIES, INCLUDING, WITHOUT LIMITATION, THE IMPLIED WARRANTIES OF
% MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE.
%
%	@(#)title.tex	5.1 (Berkeley) 7/19/90


\begin{titlepage}
\vskip 6.6em
\begin{center}
{\LARGE\Amd\ -- An Automounter}\\[.2cm]
{
%\Large \Edition\ Edition \\[.1cm]
{\DRAFT}
}
\par
\vskip 2.2em
{\large \lineskip .75em
\begin{tabular}[t]{c}
{\Large\em Jan-Simon Pendry} \\
{\tt <jsp@doc.ic.ac.uk>}\\
Department of Computing,\\
Imperial College,\\
London SW7 2BZ
\end{tabular}\par}

\vskip 1.3em

{\small May 1990\par}
{\small Printed \today}
\end{center}\par

\vskip 6em

\begin{center}
{\Large\bf Abstract}
\end{center}
{\parindent 0pt
{\bf amd}, {\em $\acute{e}$i$\acute{e}$md$\acute{\imath}$:},
a program providing an automounting service
[fr. {\em A}uto{\em M}ount {\em D}aemon]\par
{\bf automount}, {\em \"{o}'t\={o}-mownt, v.i.}
to mount a filesystem automatically
[Gr.\ {\em automatos}, self-moving $+$ \Unix\ {\bf mount}(2)]%---L.
%{\em m\={o}ns, montis}, mountain]
}
\par\mbox{}\par

An {\em automounter} maintains a cache of mounted filesystems.
Filesystems are mounted on demand when they are first referenced,
and unmounted after a period of inactivity.
\par\mbox{}\par

\Amd\ may be used as a replacement for Sun's automounter.
The choice of which filesystem to mount can be controlled dynamically
with {\em selectors}.
Selectors allow decisions of the form
``hostname is {\em this}\,'' or ``architecture is not {\em that}.''
Selectors may be combined arbitrarily.
\Amd\ also supports a variety of filesystem types,
including \NFS, \UFS\ and the novel {\em program}
filesystem.
The combination of selectors and multiple filesystem types allows
identical configuration files to be used on all machines so
reducing the administrative overhead.

\Amd\ ensures that it will not hang if a remote server
goes down.  Moreover, \amd\ can determine when a remote
server has become inaccessible and then mount replacement
filesystems as and when they become available.
\par\mbox{}\par
\Amd\ contains no proprietary source code and
has been ported to numerous flavours of \Unix.

\typeout{Lee says ``This documentation needs rewriting!''}

%\ifarticle\else
\newpage
\vspace*{30pt}
%{\large
%\parindent 0pt
%WARNING:
%%warnings go here
%}\par
\vspace{4in}
{\parindent 0pt
This document describes release ``\VERSION'' of \amd\ (also known as \AKAVERSION).\par
\vskip 1ex
Copyright \copyright\ 1989 Jan-Simon Pendry\\
Copyright \copyright\ 1989 Imperial College of Science, Technology \& Medicine\\
Copyright \copyright\ 1989 The Regents of the University of California.
\vskip 1ex
All Rights Reserved.
\vskip 1ex
\begin{flushleft}
Permission to copy this document, or any portion of it, as\\
necessary for use of this software is granted provided this\\
copyright notice and statement of permission are included.
\end{flushleft}
}
%\fi %!article
\end{titlepage}

\setheader{\vbox{\hbox to\hsize{\rightmark\hfill\thepage}\vskip 2pt\hrule}}
%\setbothheaders
%{\vbox{\hbox to\hsize{\thepage\hfill\leftmark}\vskip 2pt\hrule}}%
%{\vbox{\hbox to\hsize{\rightmark\hfill\thepage}\vskip 2pt\hrule}}

%\setfooter{\hfil\Edition\ Edition {\DRAFT}\hfil}
\setfooter{\hfil{\DRAFT}\hfil}
%\setbothfooters
%{\thepage\hfil\Edition\ Edition {\DRAFT}}%
%{\Edition\ Edition {\DRAFT}\hfil\thepage}

%\setchapterfoot{\thepage\hfil\Edition\ Edition {\DRAFT}\hfil\thepage}
%\setchapterfoot{\Edition\ Edition {\DRAFT}\hfil\thepage}
\setchapterfoot{{\DRAFT}\hfil\thepage\hfil}

\ifarticle\else
\tableofcontents
\fi
