% appendix B					% mtr

\appendix{B}{Distribution Mechanics}
The \mh5 distribution is available in two forms:
\smallskip
{\advance\leftskip by\parindent
\item{1.} Anonymous FTP\hbreak
If you can FTP to the ARPA Internet,
use anonymous FTP to the ARPAnet host UDel-Huey [10.2.0.96]
and retrieve the file portal/mh.5--tar.
This is a \pgm{tar} image of size 2.1~MB (approximately).

\item{2.} 9--track tape, 1600~bpi, \pgm{tar} format\hbreak
Otherwise, you can send \$50.00 to the address below.
This covers the cost of a magtape, handling, and shipping.
In addition,
you'll get a laser-printed hard-copy of the \MH/ documentation.
The documentation includes installation guide,
\MH/ Tutorial,
\MH/ User's Manual,
changes document (from \mh4 to \mh5),
and BBoards Manual.

\item{} If you go with this option,
be sure to include your USPS address with your check.
Checks should be made payable to
$$\hbox{Regents of the University of California}$$
It's also a good idea (though not mandatory) to send a computer mail message
to {\tx Bug-MH@UCI} when you send your check via USPS to ensure minimal
turn-around time.
The distribution address is:
$$\displayindent=\leftskip	\advance\displayindent by1.5\parindent
    \halign{\leftline{#}\cr
	Support Group\cr
	Attn: MH Distribution\cr
	Department of Information and Computer Science\cr
	University of California, Irvine\cr
	Irvine, CA  92717\cr
	\cr
	714/856--6852\cr
}$$

\item{} Sadly, if you just want the hard-copies of the documentation,
you still have to pay the \$50.00.
The \pgm{tar} image has the documentation source
(the man is in \ROFF/ format, but the rest are in \TeX\ format).
\smallskip}
\noindent
In addition,
there is some hope that \mh5, or a successor,
might be found in a future 4.x~Berkeley distribution.

Although \MH/ is not ``supported'' per se,
it does have a bug reporting address.
Normally,
the address {\tx Bug-MH@UCI} is used to report bugs and bug fixes.
There are however, 
two discussion groups which concern themselves with \MH/:
\smallskip
{\advance\leftskip by\parindent
\item{1.} MH-Users@UCI\hbreak
A discussion group for the \MH/ user community at large.
Appropriate topics include:
questions about how to use \MH/,
tips on \MH/ usage,
and exchange of \MH/ shell scripts.
All requests to be added to or deleted from this list,
along with problems,  questions and suggestions,
should be sent to {\tx MH-Users-Request@UCI}.

\item{2.} MH-Workers@UCI\hbreak
A discussion group for \MH/ maintainers and experts.
Appropriate topics include:
questions on how to configure \MH/,
tips on \MH/ configuration,
exchange of \MH/ bug reports (and fixes).
All requests to be added to or deleted from this list,
along with problems, questions and suggestions,
should be sent to {\tx MH-Workers-Request@UCI}.
\smallskip}
\noindent
The \eg{UCI} host is also known as \eg{ucivax},
so a possible \UUCP/ path might be {\tx $\tdots$!ucbvax!ucivax!bug-mh}.

Updates to \MH/ are published on the {\tx MH-Workers} list
in the form of context diffs,
and the appropriate distribution images are updated as well.
