% run this through PhD-TeX

\input dcustom
\articlestyle
\input version

\def\draftext{%
    \begingroup
	\eightpoint\sf
	Reprinted from
	{\it Proceedings, Summer Usenix Conference and Exhibition},
	Portland, Oregon,
	June, {\oldstyle 1985}%
    \endgroup
}

\input sfwmac

\def\bannerhack{$^{\bowtie}$\global\let\bannerhack=\empty}
\header
    MH.5:\\
	How to process 200 messages a day\\
	and still get some real work done\bannerhack\title
    Marshall T.~Rose\\
	Member, Research Technical Staff\\
	Northrop Research and Technology Center$^\dagger$\\
	\\
    John L.~Romine\\
	Computing Support Group\\
	Department of Information and Computer Science\\
	University of California, Irvine$^\ddagger$\author\info
\footnote{}{\hskip -\parindent $^{\bowtie}$
Alternate title: {\sl MH: Your Key to Success}.}
\footnote{}{\hskip -\parindent $^\dagger$
One~Research Park,
Palos Verdes Peninsula, CA  90274.
Telephone: 213/377--4811.\hbreak
Computer mail:
MRose\%NRTC@USC-ECL, $\tdots$!\{ucbvax!ucivax,trwrb\}!nrtc!mrose.}
\footnote{}{\hskip -\parindent $^\ddagger$
University of California at Irvine,
Irvine, CA  92717.
Telephone: 714/856--6852.\hbreak
Computer mail:
J-Romine@USC-ECL, $\tdots$!\{ucbvax,trwrb\}!ucivax!jromine.}

\centerline{\sc Abstract}			% mtr
{\rightskip=0pt\narrower
\lp
The UCI version of the Rand Message Handling System (\MH/) is discussed.
\MH/ is a powerful user agent which operates in the ARPA Internet and UUCP
environments.
In addition to the usual functions provided by similar programs,
\MH/ has several distinguishing characteristics which give the user
additional message handling capability.
In particular,
\MH/ provides mechanisms for maintaining an organized mail environment,
tailoring its behavior,
and extending its functions.

\lp
This document describes \MH/ from several perspectives.
Particular emphasis is given to:
the \MH/ user environment,
advanced features of \MH/ which have proven to be particularly useful for
sophisticated users of electronic mail,
the user interface issues in \MH/,
and the \mh5 distribution.
The paper concludes with a summary of the authors' experiences with \MH/,
and a discussion of future areas of enhancement.
\par}

\input text
\input refs
\input appendixA
\input appendixB

\printcontents
\unskip\footnote{}{\hskip -\parindent
This document (version \versiontag/)
was \TeX set \today\ with DISS.STY v\version.}
\showsummary

\bye
